Rotation and Lithium Confirmation of a 500 Parsec Halo for the Open Cluster NGC\,2516\footnote{This is paper 3 of the Cluster Difference Imaging Photometric Survey.}

### 2021-07-01 cluster_paper_review/referee_response_v0.txt

Dear Dr. Sarajedini,

We thank you for organizing the review of our paper, and also
thank the referee for constructive criticism.  We have considered
the referee's remarks carefully and revised our manuscript
accordingly.

Below, we have reproduced the relevant portions of the referee's
report, alongside our responses. A list of changes is appended to
the manuscript using the trackchanges AASTeX macros. Figures
supporting X AND Y WORDS TODO FIX THESE have also been uploaded for inspection.

Sincerely,

Luke Bouma

******************************************
REFEREE COMMENT
----------
At this stage the referee is more inclined to believe that the authors have
identified a set of young stars in that region of the sky than actual halo
member stars of NGC 2516.

RESPONSE
----------
> We agree that the stars with rotation periods similar to those of the
> Pleiades, and showing Li EWs similar to those of the Pleiades, should indeed
> be considered to have comparable ages to stars in Pleiades.
>
> Figures 4 and 7 show that the "halo" stars from KC19 and M21 overlap both
> with the Pleiades, and with stars in the core of the cluster (those from
> CG18).  They share the same gyrochronological and Li tracks, and therefore
> share the same age to within measurement precision.  The stars in this "halo"
> population are also mostly within 2 km/s of tangential velocity separation
> from the main core population (see panel in top left of Figure 6).
> Spatially, they lead and lag the core of the population, in a similar manner
> to that seen for eight other open clusters by Meingast et al 2021 (Figure 2
> of our manuscript; Figure 8 of Meingast et al 2021, Astronomy & Astrophysics,
> Volume 645, id.A84).
>
> Despite these lines of evidence for the halo stars being associated with NGC
> 2516 (same ages; same tangential velocities; spatial correlation with the
> cluster center), for the sake of caution we analyzed rotation periods of
> randomly draw field stars in this region of sky.  This exercise was described
> in Appendix C, and the results are shown in Figure 13 of the manuscript.  The
> randomly selected field stars (orange points) does not show any evidence for
> an overdensity of Pleiades-age stars.  The candidate "halo" stars selected
> kinematically by CG18, KC19, and M21 do (black points).  As a clarification
> on Figure 13, we have attached "f13_corehalosplit.png" as a supplement in the
> response, which shows the same data, but splitting NGC 2516 into halo and
> core stars.  Stars in the halo (blue) have a very different rotation-color
> distribution than randomly selected field stars (orange).
>
> In short, most of the halo stars i) have similar tangential velocites as core
> stars, ii) are coeval with other halo stars, iii) have the same age as the
> core stars, iv) are spatially correlated with the location of the core stars,
> and iv) have a different rotation-color distribution from a randomly selected
> sample of fields stars (with a large overdensity of rotators falling on the
> same sequence as the core of NGC 2516).
>
> The suggestion that we "have identified a set of young stars in that region
> of the sky [rather] than actual halo member stars of NGC 2516." is equivalent
> to asserting that the clustering techniques used by KC19 and M21 yielded a
> set of stars that is identical to the cluster core according to all age
> measurements, and is spatially and kinematically correlated with the cluster
> core, but are otherwise unrelated.  While the spatial and kinematic
> correlations are tautological -- this is what clustering methods do if they
> are working correctly -- the age measurements have no a priori reason to be
> identical between the core and halo unless the two are in fact related.

> TODO: maybe empahsize the false positive fraction of 50% here.

> Add an uncertainty on the false positive rate? Could it be higher than 50%?

******************************************
REFEREE COMMENT
----------
The authors begin by concatenating a very large collection of stars identified by
CG18 (which they term the core members), and those of KC19 and M21, the last two
having very relaxed criteria for association with NGC 2516.
At this stage, as one can see from the distributions in Fig.1, the sample is
completely bloated with field stars which now clearly have to be excised.
The question is how?

The authors justifiably look first to the HR diagram and understandably find
that that is not very helpful in rejecting non-members because almost all stars in
the extended sample lie close to the ridge line of the canonical cluster members.
(Nevertheless, this is already a basis for rejecting stars below the main sequence
and those significantly above it, including giants. The authors should consider
this.) Incidentally, the Gaia photometry for the brightest stars is likely not good
enough for fitting any isochrones of this age, so in this referee's estimation, the
cluster turnoff is perhaps still the best one can do, keeping in mind the persistence
of photometric problems for bright stars.

RESPONSE
----------
> While the HR diagram is not very helpful in rejecting non-members, it does
> indicate that the distribution of field stars in this region of
> ra/dec/parallax is different from stars in both the core and halo of the
> cluster.  In particular, the lack of a sub-giant branch or red giant branch
> (e.g., at G_BP-G_RP =1.2, M_G =3.0) suggests that the core and halo have a
> different age distribution than the field stars, and supports the suggestion
> that for the most part, the halo population has a single age.
>
> For the ~five halo stars lying significantly above the main sequence, we
> agree that these must be field interlopers.  This is noted at the end of
> Section 3.1, Paragraph 2.  The ~four halo stars, and ~four core stars lying
> well below the main sequence could be there due to poorly determined
> parallaxes, or due to blending in the Gaia BP or RP spectra with unresolved
> white dwarf companions.  We do not consider their presence below the MS to
> support their exclusion from consideration as cluster members.

******************************************
REFEREE COMMENT
----------
The authors then go on to consider the rotation periods and lithium abundances
of the extended sample, to compare them with previously-known distribution of
similarly young stars.

The conceptual problem here is that stars that deviate from the known/expected
rotational distribution for NGC 2516 are again not excised from the sample.  As
prior studies (e.g. Fritzewski+2020) have shown, there are no rotators above
the slow sequence in NGC 2516. So such stars most certainly have to be excised
from further consideration. (Some of these are even older than Praesepe, as the
authors themselves show in Fig.4.) The authors should consider whether other
regions of the color-period diagram merit similar treatment.

RESPONSE
----------

> We agree that stars showing rotation periods above the slow sequence are most
> likely not "genuine", in the sense that either i) they are not cluster
> members, or ii) they are cluster members, but with unresolved binary
> companions contaminating the rotation period measurement.  The latter
> category has precedent in the literature (e.g., Stauffer+2016, AJ 152 115,
> Section 5.1).
>
> Our point in proposing "Set A" and "Set B" is that the exact means by which
> the samples are cleaned is not incredibly important to the central result that
> the halo is the same age as the core, and therefore most likely related to
> it.
>
> Set A comprises 987 stars.  947 of these have (Bp-Rp)_0>0.4, and so make some
> degree of sense to assess gyrochronologically.  873 are below what by eye can
> be identified as the "slow sequence boundary".  While there is subjectivity
> in where this line is drawn, the big-picture result is that ~90% of the "Set
> A" stars are be "genuine" in the sense of being "in or below the slow
> sequence".  
>
> We attach two additional figures, "response_fig2.png" and
> "response_fig3.png" illustrating the point.  Cleaning out these stars to make
> a "perfectly contaminant free" list of "genuine cluster members" is not the
> aim of this work.
>
> (We note also that simply because a star is in or below the slow sequence
> does not guarantee that it is genuine).

******************************************
REFEREE COMMENT
----------
An additional concern is that the authors construct two samples A (where rotation
periods from multiple types of measurements don't agree) and B (where rotation
periods agree), but then go further with sample A (for instance Fig.6).
Had the authors gone forward with the tightest-possible sample, and had then
shown that this tight rotation-lithium sample still has a far greater extent on
sky and in the canonical parallax, proper motion, RV etc. spaces, then readers would
be far more willing to believe the authors' claims about a more extensive cluster.

RESPONSE
----------
> "response_fig2.png" and "response_fig3.png" illustrate that, for the
> rotation-color selected subsample of Set A suggested by the referee in the
> previous comment, the halo is the size claimed.
>
> The suggestion to use "Set B" for this is the suggestion to apply a selection
> function that throws away ~85% of stars.  (We suspect this is mostly because
> the SPDM periods tend to be garbage).  The halo is already sparse, so
> throwing away ~9 in 10 of its members will make it sparser.
>
> Regardless, doing this with Set B gives response_fig4.png.  The "slow
> rotating" (and below) halo stars still span a large range of space, though
> more like ~350 pc tip-to-tip rather than ~500 pc.
>
> Anyway, in Set A (N=937), 216 stars are at least 50 parsecs from the cluster
> center, and 104 are at least 100 parsecs away (23% and 11%).  In Set B
> (N=479), 94 stars are at least 50 parsecs from the cluster center, and 43 are
> at least 100 parsecs away (20% and 9%).

******************************************
REFEREE COMMENT
----------
The referee notes that the prior work of Healy & McCullough 2020, who also used
TESS data to measure a sizeable group of stars in NGC 2516 has been completely
ignored. The authors should include a discussion of the H+M2020 work in relation
to the results in this paper.

RESPONSE
----------
> Healy & McCullough 2020 (HM20) was referenced in the second paragraph of
> Section 4.1, in relation to their discussion of the contraction rate of the
> core NGC 2516, and the corresponding rate at which stars are escaping the
> core.  The form in which Healy & McCullough quantified the contraction rate
> does not readily translate (to our knowledge) to an estimate of the rate at
> which stars escape from the cluster's tidal radius, or an analogous surface.
> So, the reference was rather brief.
>
> The main result of our paper -- that the halo is the same age as the core and
> therefore most easily explained as being associated with NGC2516 -- does not
> clearly translate to the main result of HM20 -- that the stellar spins in the
> core of NGC 2516 are either isotropically oriented, or perhaps mildly
> aligned.  Therefore we did not highlight HM20 prominently in our discussion.
>
> Nonetheless, to address the general idea behind this suggestion, we have i)
> added a reference to HM20 in the introduction, when we discuss TESS and
> NGC2516, and ii) included the HM20 rotation periods in Figure 10.  The
> overall point of the figure has not changed.

******************************************
REFEREE COMMENT
----------
While on the subject, it should also be mentioned that the Introduction of the
paper is only from the galactic structure perspective, and is lacking in
introducing readers to the rotation-lithium perspective. Given that the latter
constitute the bulk of the paper, doing that would clearly help the argument
being laid out.

RESPONSE
----------
> We have added a paragraph in the introduction giving some context to this
> effect -- thank you for the suggestion.

******************************************
REFEREE COMMENT
----------
This referee came away with the general impression that the authors have likely
identified a number of young stars in a vast halo around NGC 2516, but did not
find the association with NGC 2516 itself to be particularly convincing.
Perhaps they can think about how to make their arguments tighter, and the
inclusion criterea stricter in the next iteration, to be able to better convince
the community of the association with NGC 2516.

RESPONSE
----------
> XXX of YYY stars for which we expected to detect rotation showed a detected
> rotation signature...
> In the comparison sample...

******************************************
DATA EDITOR'S COMMENT
----------
The .csv file attached with Table 1 has a large number of significant digits in
most columns. In the revised manuscript, the authors should include a new
version that has only the required precision. This document outlines the
typical uncertainties for Gaia data:

https://gea.esac.esa.int/archive/documentation/GDR2/

Fewer significant digits will result in a smaller file that will be easier to
read, manipulate, and transfer.

RESPONSE
----------
> The CSV file has been updated with an appropriate number of significant
> digits in each column; thank you for the suggestion.

### 2021-07-01 cluster_paper/maintable_header.tex
[WIP] staging for review
(not done)

               \texttt{source\_id} & 5290665204845694336 &                                                                   Gaia DR2 source identifier. \\
         \texttt{source\_id\_edr3} & 5290665204845694336 &                                                                  Gaia EDR3 source identifier. \\
                 \texttt{in\_SetA} &                True &                                 In Set $\mathcal{A}$ (LSP>0.08, P<15d, nequal==0, nclose>=1). \\
                 \texttt{in\_SetB} &                True &                                   In Set $\mathcal{B}$ (Set $\mathcal{A}$ and periods match). \\

@@

                   \texttt{period} &            0.228981 &                                                              Lomb-Scargle best period [days]. \\
                   \texttt{lspval} &            0.657826 &                                               Lomb-Scargle periodogram value for best period. \\
               \texttt{spdmperiod} &            0.228981 &                                                          Stellingwerf PDM best period [days]. \\
                  \texttt{spdmval} &            0.372779 &                                           Stellingwerf PDM periodogram value for best period. \\

@@

                       \texttt{ra} &          119.234778 &                                                               Gaia DR2 right ascension [deg]. \\
                      \texttt{dec} &          -61.015929 &                                                                   Gaia DR2 declination [deg]. \\

@@

                 \texttt{parallax} &            2.458023 &                                                                      Gaia DR2 parallax [mas]. \\
          \texttt{parallax\_error} &            0.028395 &                                                          Gaia DR2 parallax uncertainty [mas]. \\
                     \texttt{pmra} &           -4.225826 &                           Gaia DR2 proper motion $\mu_\alpha \cos \delta$ [mas$\,$yr$^{-1}$]. \\
                    \texttt{pmdec} &           10.899938 &                                       Gaia DR2 proper motion $\mu_\delta$ [mas$\,$yr$^{-1}$]. \\
       \texttt{phot\_g\_mean\_mag} &           15.778953 &                                                                       Gaia DR2 $G$ magnitude. \\
      \texttt{phot\_bp\_mean\_mag} &           16.629477 &                                                           Gaia DR2 $G_\mathrm{BP}$ magnitude. \\
      \texttt{phot\_rp\_mean\_mag} &            14.85896 &                                                           Gaia DR2 $G_\mathrm{RP}$ magnitude. \\

@@

                  \texttt{in\_M21} &               False &                                                                Star in \citet{meingast_2021}. \\
               \texttt{(Bp-Rp)\_0} &            1.636198 & Gaia $G_\mathrm{BP}$-$G_\mathrm{RP}$ color, minus $E$($G_\mathrm{BP}$-$G_\mathrm{RP}$)=0.1343 \\

@@

                     \texttt{ruwe} &            0.986349 &                                                                               Gaia EDR3 RUWE. \\
      \texttt{Li\_EW\_mA\_GaiaESO} &              -1.495 &                          Gaia-ESO Li doublet equivalent width, including the Fe blend [m\AA]. \\
\texttt{Li\_EW\_mA\_perr\_GaiaESO} &              12.407 &                                              Gaia-ESO Li doublet EW upper uncertainty [m\AA]. \\

>>>>

               \texttt{source\_id} & 5290666991552265088 &                                                                                Gaia DR2 source identifier. \\
         \texttt{source\_id\_edr3} & 5290666991552265088 &                                                                               Gaia EDR3 source identifier. \\
                 \texttt{in\_SetA} &                True &                                    In Set $\mathcal{A}$ (LSP$>$0.08, P$<$15d, nequal$==$0, nclose$\geq$1). \\
                 \texttt{in\_SetB} &                True & In Set $\mathcal{B}$ (Set $\mathcal{A}$ and below $P_\mathrm{rot}-(G_\mathrm{BP}$-$G_\mathrm{RP})_0$ cut). \\

@@

                   \texttt{period} &            7.733189 &                                                                           Lomb-Scargle best period [days]. \\
                   \texttt{lspval} &            0.505373 &                                                            Lomb-Scargle periodogram value for best period. \\

@@

                       \texttt{ra} &          119.494256 &                                                                            Gaia DR2 right ascension [deg]. \\
                      \texttt{dec} &          -61.060819 &                                                                                Gaia DR2 declination [deg]. \\

@@

                 \texttt{parallax} &            2.462602 &                                                                                   Gaia DR2 parallax [mas]. \\
          \texttt{parallax\_error} &            0.017613 &                                                                       Gaia DR2 parallax uncertainty [mas]. \\
                     \texttt{pmra} &           -5.366865 &                                        Gaia DR2 proper motion $\mu_\alpha \cos \delta$ [mas$\,$yr$^{-1}$]. \\
                    \texttt{pmdec} &           10.959816 &                                                    Gaia DR2 proper motion $\mu_\delta$ [mas$\,$yr$^{-1}$]. \\
       \texttt{phot\_g\_mean\_mag} &           14.592845 &                                                                                    Gaia DR2 $G$ magnitude. \\
      \texttt{phot\_bp\_mean\_mag} &           15.189163 &                                                                        Gaia DR2 $G_\mathrm{BP}$ magnitude. \\
      \texttt{phot\_rp\_mean\_mag} &           13.867151 &                                                                        Gaia DR2 $G_\mathrm{RP}$ magnitude. \\

@@

                  \texttt{in\_M21} &                True &                                                                             Star in \citet{meingast_2021}. \\
               \texttt{(Bp-Rp)\_0} &            1.187693 &              Gaia $G_\mathrm{BP}$-$G_\mathrm{RP}$ color, minus $E$($G_\mathrm{BP}$-$G_\mathrm{RP}$)=0.1343 \\

@@

                     \texttt{ruwe} &            0.954835 &                                                                                            Gaia EDR3 RUWE. \\
      \texttt{Li\_EW\_mA\_GaiaESO} &             144.259 &                                       Gaia-ESO Li doublet equivalent width, including the Fe blend [m\AA]. \\
\texttt{Li\_EW\_mA\_perr\_GaiaESO} &               6.301 &                                                           Gaia-ESO Li doublet EW upper uncertainty [m\AA]. \\
